\section*{Declarations}

\subsection*{Ethics approval and consent to participate}
Not applicable --- all analyses use previously published, de-identified
public datasets.

\subsection*{Consent for publication}
Not applicable.

\subsection*{Availability of data and materials}
All raw data and processed intermediate files are available as
described in the Methods section (Data availability).
Supplementary tables accompanying this manuscript:
S1 (cell-cycle ablation permutation $p$-values),
S2 (EMT-filter Mapper gene connectivity),
S2$'$ (PC1-filter Mapper gene connectivity),
S3 (Mapper parameter sweep),
S4 (PDX gene attribution),
S5 (Max $H_1$ vs principal component count),
S6 (functional persistence summaries: max $H_1$, total persistence,
persistence entropy across ten cohorts),
S7 (subsample variance: mean $\pm$ SD of max $H_1$ across five random
seeds for four cohorts),
S8 (Mapper parameter-sweep Spearman rank correlations),
S9 (variance-normalized Mapper gene connectivity),
S10 (Harmony batch-correction raw vs.~corrected max $H_1$ comparison),
S11 (Maynard 2020 full EGFR-mutant cohort per-patient max $H_1$ and within-patient trajectory statistics (14 patients, 3 matched pre/post pairs)),
S-stringent-CC (stringent cell-cycle regression ratios; reproduced in
the lower block of Table~\ref{tab:cc_controls}).

\subsection*{Competing interests}
The authors declare that they have no competing interests.

\subsection*{Funding}
This work received no specific external grant funding and was carried
out using the authors' institutional computing resources at Koo
Foundation Sun Yat-Sen Cancer Center (H.-T.L.) and the Center for
Computational and Systems Biology, National Taiwan University
(Y.-H.T.). \texttt{<grant numbers TBD>}.

\subsection*{Authors' contributions}
H.-T.L. (first author) and Y.-H.T. (corresponding author) jointly
conceived the study and designed the analytical strategy. H.-T.L.
implemented the pipeline, performed all analyses, produced all
figures, and wrote the original draft (CRediT: Conceptualisation,
Methodology, Software, Formal analysis, Investigation, Data curation,
Writing -- original draft, Visualisation). Y.-H.T. supervised the
project, contributed methodological and validation-strategy review,
and revised the manuscript (CRediT: Conceptualisation, Methodology,
Supervision, Writing -- review \& editing, Project administration).
Both authors read and approved the final manuscript.
