\section{Results}

\subsection*{Erlotinib-treated PC9 cells display robust $H_1$ persistence (max $H_1 = 3.78$ at day 9)}

We first asked whether drug-treated PC9 cells (harboring EGFR exon 19
deletion E746-A750del) exhibit non-trivial $H_1$ features consistent
with cyclic state-space structure. We analyzed
GSE134839 \citep{aissa2021}, a Drop-seq time-series of PC9 cells treated
with \SI{2}{\micro\molar} erlotinib at days 0, 1, 2, 4, 9, and 11
(\cref{fig:study_overview}). After standard QC and preprocessing
(Methods), 2{,}942 cells across six timepoints were projected into the
top-30 principal components and used to compute the persistence diagrams
via the Vietoris--Rips filtration.

All drug-treated timepoints showed non-zero $H_1$ persistence, with
maximum values ranging from 1.08 (D11) to 1.77 (D9)
(\cref{fig:discovery_ph}, \cref{tab:per_timepoint}). Persistence barcodes
showed an increased number of long $H_1$ bars at intermediate and late
timepoints. Pairwise Wasserstein distances between persistence diagrams
revealed that D0 is separated from all drug-treated timepoints in persistence-diagram space (Wasserstein distance $W = 213$--$265$ between D0 and treated timepoints vs.~$W = 26$--$66$ among treated timepoints --- roughly an order of magnitude larger), indicating that the distribution of topological features differs most strongly between untreated and treated conditions (\cref{fig:discovery_ph}).

Per-timepoint sample sizes (200--379 cells) left individual
permutation tests underpowered (closest D9 $p = 0.098$, D2
$p = 0.126$, near the peak of persister emergence reported by Aissa
et al.); the larger validation cohorts below recover $p < 0.01$
(\cref{tab:validation}).

\subsection*{Loops persist after Tirosh cell-cycle ablation across five cohorts (ratio $0.71$--$1.11$)}

To rule out cell cycle oscillation as a confounder, we removed 39 of 97
Tirosh et al.\ cell-cycle genes \citep{tirosh2016} present in our HVG
set, re-scaled, and re-computed PCA and persistent homology. The ratio
of $H_1$ after to before cell-cycle ablation ranged from 0.71 to 1.11
across all six timepoints (mean $0.90 \pm 0.14$), with D11 actually
\emph{increasing} after cycle-gene removal (ratio $= 1.11$) ---
indicating that the observed cyclic topology is not driven by
cell-cycle genes (\cref{fig:discovery_ph}, \cref{tab:cc_controls}). This
ratio-based test is the standard control in the field and has been
interpreted as evidence against cell-cycle artifacts.

(Per-timepoint nulls on the ablated discovery cohort were
$p = 0.11$--$0.58$, expected at the small $n$ further degraded by
removing 39 HVGs; larger validation cohorts retain $p < 0.01$;
Table~S1.)

\subsection*{Stringent S/G2M score regression preserves loops in all five cohorts (ratio $1.07$--$4.05$)}

Stringent \texttt{sc.pp.regress\_out(S\_score, G2M\_score)} on
1{,}000-cell subsamples preserves loops across five cohorts
(post/pre ratio 1.07--4.05, mean 1.97, every value $> 1$;
\cref{tab:cc_controls}; full discussion in Supplementary
Methods~S13). The osimertinib cohort additionally contains zero
Tirosh genes in its HVG set, ruling out Tirosh-specific contribution
to that cohort by absence.

\subsection*{Osimertinib drives a 2.7$\times$ monotonic max $H_1$ increase over 14 days}

On GSE150949 \citep{oren2021}, a PC9 lineage-tracing osimertinib
time-series (56{,}419 cells subsampled to 5{,}000; D0/D3/D7/D14;
4{,}999 post-QC), max $H_1$ rose monotonically $1.41 \to 1.82 \to
2.67 \to 3.78$ (2.7$\times$; \cref{fig:validation}), with
permutation tests confirming $p < 0.01$ at D7 and D14
(\cref{tab:validation}).

\subsection*{PDX xenografts replicate the osimertinib-driven cycle (YU-006: 2.2$\times$, YU-003: 1.4$\times$)}

In GSE243562 \citep{wurtz2024} --- two EGFR-mutant PDX models
(YU-003, YU-006) under matched untreated and osimertinib-residual
conditions, 18{,}874 cells subsampled to 1{,}500 per group --- both
models showed an increase in max $H_1$ under residual disease:
YU-003 $2.79 \to 3.85$ ($1.4\times$, $p < 0.01$); YU-006
$2.81 \to 6.08$ ($2.2\times$, $p < 0.01$) (\cref{fig:validation},
\cref{tab:validation}). YU-006 matches the cell-line D14 effect;
YU-003 is more modest but directionally consistent. This replicates
the cell-line finding in an \emph{in vivo} system with stromal,
vascular, and host-immune context.

\subsection*{Max $H_1$ in drug-treated cells approaches patient-tumor levels}

To establish a topological reference for untreated human lung
adenocarcinoma, we analysed GSE131907 \citep{kim2020} --- a
208{,}506-cell LUAD atlas across 44 patients --- subsampling 5{,}000
primary-tumour epithelial cells (tLung), then 1{,}500 for VR
computation. Treatment-naive patient LUAD shows max $H_1 = 4.77$,
substantially higher than treatment-naive PC9 cell line (1.41) or
PDX (2.79, 2.81) baselines (\cref{fig:master}). The Kim atlas is
mixed-driver (EGFR-, KRAS- and other-mutant tumours; not
genotype-selected), so this is a generic LUAD reference rather than
an EGFR-specific baseline.

Drug treatment moves cell-line and PDX topology from the
cell-line-like regime ($H_1 \approx 1$--$3$) toward the mixed-driver
patient range ($H_1 \approx 4$--$5$; \cref{fig:master}, closest at osi
D14 and YU-006 residual) --- a qualitative cross-system comparison
confounded by platform and cell count; the load-bearing claim is the
within-system monotonic increase
(\cref{fig:validation}).

\subsection*{Timepoint-label permutation null confirms drug-specific signal}

The gene-label null tests for non-random topology but not for drug
specificity, so we ran a timepoint-label null on osimertinib D14 vs.\
D0 (drug-exposure annotation shuffled across the pooled 2{,}499 cells;
$H_1$ recomputed in each permutation). Over 50 permutations, the
observed $\Delta H_1 = 2.07$ exceeded the 95th percentile (1.18) in
0/50 trials, giving $p < 0.02$ (Phipson--Smyth $+1$ correction) ---
direct evidence that $H_1$ depends on drug exposure.

\subsection*{EMT genes dominate loop-associated transcripts in both PC9 and PDX systems}

What do the $H_1$ loops represent biologically? To identify the
transcriptional programs underlying loop formation, we constructed
Mapper graphs on the analyzed cell line data using EMT score as filter
function (Methods). The Mapper graph contained 20 nodes and 16 edges;
gene-level connectivity scores quantify how strongly each gene varies
between Mapper nodes versus globally.

In the cell-line analysis, top-connected genes were canonical EMT
markers (\emph{TPM1}, \emph{KRT8}, \emph{KRT19}); MSigDB Hallmark EMT
ranked first in top-100 pathway enrichment. The PDX analysis
recovered a broader mesenchymal/secretory signature (\emph{S100A4},
\emph{LGALS3}, \emph{COL1A1}) --- 3/33 vs.\ 8/33 Hallmark EMT genes
in the cell-line top 100 --- consistent with an EMT-like plastic
state realised through a partially shared repertoire. Eleven genes
overlap between the two top-100 lists (including \emph{KRT19},
\emph{LGALS3}); hypergeometric $p = 1.25 \times 10^{-11}$ against
the 13{,}955-gene background of features expressed in both analyses,
$p = 2.1 \times 10^{-4}$ against the more conservative
HVG-intersection background (~3{,}000 genes). Differential expression
recovered EMT-related signatures in both systems with
system-specific second-order pathways (cholesterol homeostasis
downregulated in the cell line; estrogen-response upregulated in PDX,
a pathway known to interact with EMT via estrogen-receptor--mediated
transcription in NSCLC \citep{siegfried2014}). Full gene-level lists
are in Supplementary Tables~S2, S2$'$, and S4.

\paragraph{Robustness of gene attribution.}
Two sensitivity checks support the EMT readout. With PC1 as an
unbiased Mapper filter, only 4/33 Hallmark EMT genes appear in the
top 100 (vs.~8/33 with the EMT filter) and cell-cycle markers
(\emph{MKI67}, \emph{TOP2A}, \emph{HMGB2}) co-rank with EMT markers,
so the loops reflect a multi-program state-space rather than a pure
EMT signature; PC1$\leftrightarrow$EMT top-100 overlap is 53/100.
Variance-normalised connectivity (dividing each gene's contribution
by its global SD) raises Hallmark-EMT hits to 13/33 in the top 100
and drops housekeeping genes out, arguing the enrichment is not an
expression-magnitude artefact. Full tables: Supplementary S2 (EMT
filter), S2$'$ (PC1 filter), S9 (variance-normalised).

\subsection*{The framework is modular, reproducible, and generalises beyond drug tolerance}

The pipeline is fifteen independent modules coordinated by a single
\texttt{make pipeline} target (Methods; Supplementary Methods~S8).
The same framework ran without modification on Drop-seq and 10x
cell-line data (6{,}508--56{,}419 raw cells), 10x PDX data with
automated mouse-gene filtering (22{,}032 cells), and a 208{,}506-cell
patient atlas. Applying it to a new dataset needs only (i) a QC'd and
normalised \texttt{AnnData} object with a sample/timepoint column,
(ii) a Mapper filter (PC1, EMT score, or diffusion component~1 are
shipped as defaults), and (iii) a permutation-test subsampling budget
(1{,}000 cells per group is sufficient at $p < 0.05$). We anticipate
the framework will generalise to other scRNA-seq perturbation contexts
including chemotherapy, immunotherapy escape, and stress-induced
dedifferentiation. Resampling stability (CV 8.9--23.5\% with ordering
preserved), replication under alternative persistence-diagram summaries
(total persistence and persistence entropy), and Mapper-parameter
robustness (mean Spearman $\rho = 0.994$ across a $4 \times 5$
parameter sweep) are reported in Supplementary Methods~S10--S12.

\subsection*{The monotonic trend survives Harmony batch correction}

To rule out that the observed $H_1$ growth reflects batch or platform
effects rather than intrinsic biology, we applied Harmony batch
correction \citep{korsunsky2019} to the top-50 PCA components with
dataset-appropriate batch keys (full parameters in Supplementary
Methods~S7). Max $H_1$ was then recomputed on the Harmony-corrected
embeddings; Figure~\ref{fig:harmony} summarises the comparison.

The osimertinib monotonic trend survives Harmony correction
($1.35 \to 3.94$; $2.9\times$, vs.~$1.71 \to 3.78$ raw); the YU-006
untreated$\to$residual difference is preserved ($2.59 \to 5.19$
Harmony vs.~$2.36 \to 5.01$ raw); YU-003 becomes more modest but
directionally consistent; the Kim LUAD naive baseline is robust
(Harmony $4.52$ vs.~raw $4.77$). Full values: Supplementary Table~S10.
The $H_1$ growth across drug exposure therefore reflects biology
rather than within-dataset batch structure.

\subsection*{In 14 EGFR-mutant NSCLC patients, max $H_1$ grew monotonically TN $\to$ PER $\to$ PD ($5.37 \to 7.13 \to 8.05$)}

To directly test the framework on clinical tissue, we analysed the
full EGFR-mutant subset of \citet{maynard2020}: 13{,}244 Smart-seq2
cells across 14 patients with biopsies at three treatment stages ---
treatment-naive (TN, 3{,}287 cells), persister/residual disease (PER,
4{,}893 cells), and progressive disease (PD, 5{,}064 cells). Across
1{,}500-cell subsamples per stage, pooled max $H_1$ grew monotonically
$5.37 \to 7.13 \to 8.05$ (TN-to-PD $1.50\times$;
\cref{fig:patient}a).

\paragraph{Per-patient matched trajectories.}
Three patients had matched TN + post-treatment biopsies: TH218
($1.22 \to 1.65$, $1.35\times$), TH226 ($2.11 \to 6.86$, $3.25\times$;
$\to 5.01$ PD, $2.37\times$), and AZ003 ($5.32 \to 8.50$, $1.60\times$;
\cref{fig:patient}a). Post-treatment max $H_1$ exceeded TN in 3/3
matched pairs (median $1.60\times$, range $1.35$--$3.25\times$;
one-sided sign test $p = 0.125$; underpowered at $n = 3$ but
directionally consistent; Supplementary Table~S11). Raw max $H_1$
spans $\sim 1.2$--$7$ across patients at any stage --- inter-patient
biological and technical heterogeneity --- so the within-patient
trajectory (always up) is a stronger signal than absolute magnitude,
supporting use of max $H_1$ as a \emph{relative-change} readout.

This patient-cohort result is the most clinically proximal evidence
we provide: the same topological signature that rises under drug in
PC9 and PDX also rises in real EGFR-mutant NSCLC patient biopsies, in
a hypothesis-generating three-of-three matched design.
